A Meta-analysis of Longevity Estimates of Mosquito Vectors of Disease: Supplementary Material

### 2018-11-23 MS.tex
Changed bibliography to thesis version and started adding references

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\begin{document}

@@

Mosquitoes are responsible for more human deaths than any other
organism, yet we still know relatively little about their ecology.
Mosquito lifespan is a key determinant of transmission strength for the
diseases they vector, but the field experiments used to determine this
quantity -- mark-release-recapture (MRR) studies and wild-caught
dissection of female mosquitoes -- produce estimates with high
uncertainty. In this paper, we use Bayesian hierarchical models to
analyse a previously-published database of 232 MRR experiments and
compile then analyse a database of 131 dissection studies to produce the
first ever species- and genus-level estimates of mosquito lifespan. Due
to the assumptions required to analyse the field data, we term our
estimates lower bounds on lifespan (LBL). Notably, for the major African
malaria vector \emph{Anopheles gambiae s.l.}, we estimate LBLs of 4.4
days (median estimate; 25\%-75\% CI: 3.8-5.1 days for unfed female
mosquitoes from the MRR analysis) and 7.6 days (median estimate;
25\%-75\% CI: 5.2-11.0 days from the dissection analysis); and an LBL of
4.2 days (median estimate; 25\%-75\%
CI: 3.6-4.8 days, only present in the MRR database) for the
predominantly East-African vector \emph{A. funestus s.l}. We estimate
LBLs of 6.2 days (median estimate; 25\%-75\% CI: 4.5-8.5 days from the
MRR analysis) and 4.3 days (median estimate; 25\%-75\% CI: 3.5-5.1 days
from the dissection analysis) for \emph{Aedes aegypti}; and 11.6 days
(median estimate; 25\%-75\% CI: 10.0-13.7 days from the MRR analysis)
for \emph{Ae.} \emph{albopictus} -- the predominant vectors of dengue
fever, chikungunya and Zika. Our estimates indicate that there is
significant variation in lifespan across species, with most variation
explained by differences between genera. In correspondence with
laboratory studies, we estimate that female mosquitoes outlive males by
0.9 days on average (median estimate; 25\%-75\% CI: 0.3-1.6 days). We
fit models incorporating mosquito senescence to the data, which allows
us to assess evidence for age-dependent mortality in mosquitoes across
different species. We determine that 8 of 33 species included in the MRR
database indicated evidence for senescence, versus only 2 of 25 species
from the dissection database. Our analysis applies a common framework to
the analysis of databases of MRR and dissection-based experiments,
allowing us to produce robust estimates of lower bounds on lifespan. It
also enables us to critically appraise each field method, which
highlights a need for alternative field methods for measuring this
important mosquito characteristic.

@@

mosquitoes, mortality, senescence, vector-borne disease, Bayesian,
hierarchical model.

@@

Some of the most important infectious diseases afflicting humans are
transmitted by mosquitoes, including pathogens such as the causative
agent of malaria that have been associated with humans throughout our
evolutionary history as well recently emergent infections such as the
Zika virus. Most mosquito species have a ``gonotrophic cycle'' involving
successive episodes of vertebrate blood feeding, egg maturation and
oviposition. In order for a mosquito to transmit a pathogen it must feed
on an infectious person and live long enough to complete at least one
gonotrophic cycle and feed on an uninfected and susceptible individual.
Adult lifespan is thus a critical determinant of the ability of a
mosquito population to allow the persistence of an indirectly
transmitted infection. Lifespan can of course be straightforwardly
assessed in the laboratory, but it is generally accepted that
measurements under relatively benign laboratory conditions are likely to
have limited relevance in the field, and much effort has been directed
at estimating this parameter in the vector's natural environment. Most
work has focused on assessing average daily mortality rates, and the
simplest assumption is that these do not vary with mosquito age -- in
this case longevity is simply the reciprocal of mortality. Testing this
assumption and discovering whether mosquitoes senesce or show other
types of age-dependent mortality has also been studied in the field.

@@

\section{References}\label{references}

ADDIN EN.REFLISTX

\section{Figure Legends}\label{figure-legends}

Figure 1:

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage[citestyle=authoryear,natbib=true,backend=bibtex,giveninits=true]{biblatex}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\bibliography{Malaria}
\begin{document}

@@

Mosquitoes are responsible for more human deaths than any other
organism, yet we still know relatively little about their ecology.
Mosquito lifespan is a key determinant of transmission strength for the
diseases they vector, but the field experiments used to determine this
quantity -- mark-release-recapture (MRR) studies and wild-caught
dissection of female mosquitoes -- produce estimates with high
uncertainty. In this paper, we use Bayesian hierarchical models to
analyse a previously-published database of 232 MRR experiments and
compile then analyse a database of 131 dissection studies to produce the
first ever species- and genus-level estimates of mosquito lifespan. Due
to the assumptions required to analyse the field data, we term our
estimates lower bounds on lifespan (LBL). Notably, for the major African
malaria vector \emph{Anopheles gambiae s.l.}, we estimate LBLs of 4.4
days (median estimate; 25\%-75\% CI: 3.8-5.1 days for unfed female
mosquitoes from the MRR analysis) and 7.6 days (median estimate;
25\%-75\% CI: 5.2-11.0 days from the dissection analysis); and an LBL of
4.2 days (median estimate; 25\%-75\%
CI: 3.6-4.8 days, only present in the MRR database) for the
predominantly East-African vector \emph{A. funestus s.l}. We estimate
LBLs of 6.2 days (median estimate; 25\%-75\% CI: 4.5-8.5 days from the
MRR analysis) and 4.3 days (median estimate; 25\%-75\% CI: 3.5-5.1 days
from the dissection analysis) for \emph{Aedes aegypti}; and 11.6 days
(median estimate; 25\%-75\% CI: 10.0-13.7 days from the MRR analysis)
for \emph{Ae.} \emph{albopictus} -- the predominant vectors of dengue
fever, chikungunya and Zika. Our estimates indicate that there is
significant variation in lifespan across species, with most variation
explained by differences between genera. In correspondence with
laboratory studies, we estimate that female mosquitoes outlive males by
0.9 days on average (median estimate; 25\%-75\% CI: 0.3-1.6 days). We
fit models incorporating mosquito senescence to the data, which allows
us to assess evidence for age-dependent mortality in mosquitoes across
different species. We determine that 8 of 33 species included in the MRR
database indicated evidence for senescence, versus only 2 of 25 species
from the dissection database. Our analysis applies a common framework to
the analysis of databases of MRR and dissection-based experiments,
allowing us to produce robust estimates of lower bounds on lifespan. It
also enables us to critically appraise each field method, highlighting a need for alternative field methods for measuring this
important mosquito characteristic.

@@

mosquitoes, mortality, senescence, mark-release-recapture, vector-borne disease, Bayesian, hierarchical model.

@@

Some of the most important infectious diseases afflicting humans are
transmitted by mosquitoes \citep{gates2014}, including pathogens such as the causative
agent of malaria that have been associated with humans throughout our
evolutionary history \citep{carter2002evolutionary} as well recently emergent infections such as the
Zika virus \citep{world2016statement}. Most mosquito species have a ``gonotrophic cycle'' involving
successive episodes of vertebrate blood feeding, egg maturation and
oviposition \citep{silver2007mosquito}. In order for a mosquito to transmit a pathogen it must feed
on an infectious person and live long enough to complete at least one
gonotrophic cycle and feed on an uninfected and susceptible individual.
Adult lifespan is thus a critical determinant of the ability of a
mosquito population to allow the persistence of an indirectly
transmitted infection \citep{macdonald1957epidemiology}. Lifespan can of course be straightforwardly
assessed in the laboratory, but it is generally accepted that
measurements under relatively benign laboratory conditions are likely to
have limited relevance in the field, and much effort has been directed
at estimating this parameter in the vector's natural environment \citep{clements1981analysis,guerra2014global}. Most
work has focused on assessing average daily mortality rates, and the
simplest assumption is that these do not vary with mosquito age -- in
this case longevity is simply the reciprocal of mortality. Testing this
assumption and discovering whether mosquitoes senesce or show other
types of age-dependent mortality has also been studied in the field \citep{clements1981analysis,harrington2008age,hugo2014adult}.

@@
